---
title: "Exploration of DO and hypoxia graphs"
date: today
prefer-html: true
format: 
    pdf:
        echo: false
        toc: true
    html:
        toc: true
        toc-location: left
        code-fold: true
        embed-resources: true
warning: false
message: false
error: true
---

```{r}
library(tidyverse)
```

# Daily timing  

Commented out the chunk below in favor of saving its output and re-loading.  

```{r}
load(here::here("data_wq", "wq_daily.RDS"))
```


```{r}
# # find and read all wq files
# path <- here::here("data_wq", "00b_QAQCd_by_stn")
# stns_wq <- stringr::str_sub(dir(path, pattern = "wq_qc.RData$"), end = -10)
# dat_wq <- list()
# 
# for(i in seq_along(stns_wq)){
#     stat <- stns_wq[i]
#     file_in <- here::here(path, paste0(stat, "_qc.RData"))
#     dat <- get(load(file_in))
# 
#     dat <- dat |>
#         mutate(date = lubridate::as_date(datetimestamp)) |>
#         summarize(.by = date,
#                   doLT2_hrs = sum(do_mgl <= 2, na.rm = TRUE)/4,
#                   doLT5_hrs = sum(do_mgl <= 2, na.rm = TRUE)/4,
#                   across(c(temp:turb), function(x) round(mean(x, na.rm = TRUE), 2)))
# 
#     dat$station <- stat
# 
#     dat_wq[[i]] <- dat
# 
# }
# 
# wq_daily <- bind_rows(dat_wq) |>
#     select(station, everything())
# 
# save(wq_daily, file = here::here("data_wq", "wq_daily.RDS"),
#         compress = "xz")
```

The `wq_daily` data frame has total number of hours that DO was less than 2 and 5, and all other values are daily means.  

## general  

```{r}
hist(wq_daily$doLT2_hrs,
     breaks = 50,
     main = "Hours per day that are hypoxic",
     xlab = "daily hours where DO < 2 mg/L")

hist(wq_daily$doLT2_hrs,
     breaks = 50,
     main = "Hours per day that are hypoxic (trimmed y-axis)",
     xlab = "daily hours where DO < 2 mg/L",
     ylim = c(0, 10000))
```

How many days (in the entire QC'd dataset) experience hypoxia? 

```{r}
cat(prettyNum(sum(wq_daily$doLT2_hrs > 0), big.mark = ","))
```

Out of how many days represented?  

```{r}
cat(prettyNum(sum(!is.na(wq_daily$doLT2_hrs)), big.mark = ","))
```



How many experience it for more than 2 hours?  

```{r}
cat(prettyNum(sum(wq_daily$doLT2_hrs >= 2), big.mark = ","))
```

So more than half of our days with hypoxia experience it for at least 2 hours.  


How many stations have at least a single day in the record with at least 2 hours of hypoxia?  

```{r}
wq_daily |> 
    filter(doLT2_hrs >= 2) |> 
    select(station) |> 
    distinct() |> 
    nrow()
```

Out of how many stations in the dataset?  

```{r}
length(unique(wq_daily$station))
```

So most stations experience hypoxia at SOME point in the data record. The question is, where does it happen a lot. And how do we want to define "a lot".  

## number days per year and month....  

```{r}
wq_monthly <- wq_daily |> 
    mutate(year = lubridate::year(date),
           month = lubridate::month(date)) |> 
    summarize(.by = c(station, year, month),
              nDaysAnyHypoxia = sum(doLT2_hrs > 0),
              nDays2HrsHypoxia = sum(doLT2_hrs >= 2),
              Total_hours_LT2 = sum(doLT2_hrs, na.rm = TRUE),
              Total_hours_LT5 = sum(doLT5_hrs, na.rm = TRUE))

wq_annual <- wq_daily |> 
    mutate(year = lubridate::year(date)) |> 
    summarize(.by = c(station, year),
              nDaysAnyHypoxia = sum(doLT2_hrs > 0),
              nDays2HrsHypoxia = sum(doLT2_hrs >= 2),
              Total_hours_LT2 = sum(doLT2_hrs, na.rm = TRUE),
              Total_hours_LT5 = sum(doLT5_hrs, na.rm = TRUE))
```

```{r}
p <- ggplot(wq_annual,
       aes(x = year,
           y = nDays2HrsHypoxia,
           group = station)) +
    geom_line(col = "gray") +
    theme_bw()

p

# plotly::ggplotly(p)
```

```{r}
hist(wq_annual$Total_hours_LT2, breaks = 30)
hist(wq_annual$Total_hours_LT2, breaks = 30, ylim = c(0, 200))
```

# Map 1: Where is hypoxia worse than normal  

## station 'normals'  

For the table below, the number of hours where DO was less than 2 ("...LT2") and less than 5 ("...LT5") were calculated for each year at each station by counting the number of readings less than each threshold and dividing by 4. The numbers in the table are the means and medians of these yearly totals by station.  


Missing data are simply omitted from calculations; I am hoping that a similar amount of data is missing from each station each year, so we don't introduce bias and any variability is just part of the general noise.  

```{r}
norms <- wq_annual |> 
    summarize(.by = station,
              median_days2hrs = median(nDays2HrsHypoxia, na.rm = TRUE),
              mean_days2hrs = round(mean(nDays2HrsHypoxia, na.rm = TRUE), 3),
              q3_days2hrs = quantile(nDays2HrsHypoxia, prob = 0.75, na.rm = TRUE),
              median_hours_LT2 = median(Total_hours_LT2, na.rm = TRUE),
              mean_hours_LT2 = round(mean(Total_hours_LT2, na.rm = TRUE), 3),
              median_hours_LT5 = median(Total_hours_LT5, na.rm = TRUE),
              mean_hours_LT5 = round(mean(Total_hours_LT5, na.rm = TRUE), 3),
              sd_hours_LT2 = sd(Total_hours_LT2, na.rm = TRUE),
              sd_hours_LT5 = sd(Total_hours_LT5, na.rm = TRUE))
```

```{r}
norms |> 
    arrange(station) |> 
    select(station, median_hours_LT2, mean_hours_LT2, median_hours_LT5, mean_hours_LT5) |> 
    knitr::kable()
```

## annual anomalies  

Number of hours LT2 or 5, minus the median of each, and divided by the standard deviation of annual total hours (to put everything on a scale with sd 1). Not dividing by sd made the scales very extreme (skim through the table above for an idea of scale - some stations normally have 1-2 thousand hours of hypoxia per year).    

```{r}
anom_annual <- left_join(wq_annual, norms, 
                         by = "station",
                         relationship = "many-to-one") |> 
    mutate(anom_LT2 = Total_hours_LT2 - median_hours_LT2,
           anom_LT5 = Total_hours_LT5 - median_hours_LT5,
           reserve = substr(station, 1, 3),
           anom_LT2_scaled = anom_LT2 / sd_hours_LT2,
           anom_LT5_scaled = anom_LT5 / sd_hours_LT5) |> 
    select(reserve, station, year, 
           anom_LT2, anom_LT5,
           anom_LT2_scaled, anom_LT5_scaled,
           Total_hours_LT2, median_hours_LT2,
           Total_hours_LT5, median_hours_LT5)
    
```

EXTREMELY ROUGH GRAPHS  

These are saved as pdfs so you can view them fulll-screen.  

```{r}
#| fig-width: 8
#| fig-height: 6


ggplot(anom_annual,
       aes(x = year,
           y = anom_LT2_scaled,
           group = station)) +
    geom_line() +
    facet_wrap(~reserve, nrow = 6) +
    theme_bw()

ggsave(here::here("figs", "20241207-03-normalizedHypox.pdf"),
       width = 16, height = 8, units = "in")
```


Let's highlight years that are >= 1.5 sds higher than the median, and see if we can reasonably call these "bad" years.  

# Update - use 2 sds

```{r}
#| fig-width: 8
#| fig-height: 6


def_bad <- 1.5

anom2 <- anom_annual |> 
    filter(anom_LT2_scaled >= def_bad)
ggplot(anom_annual,
       aes(x = year,
           y = anom_LT2_scaled,
           group = station)) +
    geom_line() +
    geom_point(data = anom2,
               aes(x = year, y = anom_LT2_scaled),
               col = "red") +
    facet_wrap(~reserve, nrow = 6) +
    theme_bw()

ggsave(here::here("figs", "20241207-04-normalizedHypox_BadHighlighted.pdf"),
       width = 16, height = 8, units = "in")
```

Maybe a map has points on it for a reserve that had a bad year, and the points are sized by the number of *stations* having a bad year - bigger point when it's all 4 stations (like rookery bay in 2017 after Irma).  

```{r}
anom3 <- anom_annual |> 
    summarize(.by = c(year, reserve),
              nStations_LT2BadYear = sum(anom_LT2_scaled >= def_bad, na.rm = TRUE),
              nStations_LT5BadYear = sum(anom_LT5_scaled >= def_bad, na.rm = TRUE),
              nStationsReporting = length(unique(station)))
hypox_by_reserve <- anom3
save(hypox_by_reserve, file = here::here("data_wq", "hypox_by_reserve_annual.RDS"),
     compress = "xz")

anom4 <- anom3 |> 
    summarize(.by = year,
              nStations_LT2BadYear = sum(nStations_LT2BadYear, na.rm = TRUE),
              nStations_LT5BadYear = sum(nStations_LT5BadYear, na.rm = TRUE))
```


We've talked some about the difference between 2 and 5, and at least on the normalized scale, the numbers are exactly the same:  

```{r}
ggplot(anom4,
       aes(x = year)) +
    geom_line(aes(y = nStations_LT2BadYear,
                  col = "Stations with extreme DO < 2"), 
              linewidth = 2) +
    geom_line(aes(y = nStations_LT5BadYear,
                  col = "Stations with extreme DO < 5"),
              linewidth = 1) +
    theme_bw()
```

The median ("normal") amount of stations experiencing a "bad" hypoxic year is:  

```{r}
median(anom4$nStations_LT2BadYear)
```

There are only 5 years in this dataset where 15 or more stations had a "bad" year.  


```{r}
hist(norms$median_days2hrs, breaks = 30,
     ylim = c(0, 15))

hist(norms$mean_days2hrs, breaks = 30,
     ylim = c(0, 15))
```

```{r}
norms_annual <- wq_annual |> 
    summarize(.by = year,
              median = median(nDays2HrsHypoxia, na.rm = TRUE),
              mean = mean(nDays2HrsHypoxia, na.rm = TRUE),
              q3 = quantile(nDays2HrsHypoxia, probs = 0.75, na.rm = TRUE),
              pct90 = quantile(nDays2HrsHypoxia, probs = 0.9, na.rm = TRUE),
              pct95 = quantile(nDays2HrsHypoxia, probs = 0.95, na.rm = TRUE))
```


```{r}
ggplot() +
    geom_boxplot(data = wq_annual,
                 aes(x = year,
                     y = nDays2HrsHypoxia,
                     group = year)) +
    geom_line(data = norms_annual,
              aes(x = year, y = q3),
              col = "blue",
              linewidth = 0.8)+
    geom_line(data = norms_annual,
              aes(x = year, y = pct90),
              col = "blue",
              linewidth = 0.8)+
    geom_line(data = norms_annual,
              aes(x = year, y = pct95),
              col = "blue",
              linewidth = 0.8) +
    theme_bw() +
    labs(y = "Number of days hypoxic, per station")
```


# General - Monthly proportions  

## summarized   

-  how many days each month have >2 hrs of <2 mg/L  
-  how many days in summer (May - September) have >2 hrs of <2 mg/L



```{r}
wq_monthly <- read.csv(here::here("data_wq", 
                                  "00c_QAQCd_monthly_byType",
                                  "SWMP_monthlyWQ.csv")) |> 
    mutate(doLT2 = doLessThan2_total / doLessThan2_nValid,
           doLT5 = doLessThan5_total / doLessThan5_nValid)

trnds <- read.csv(here::here("data_wq",
                             "02_calculated_long-term-trends",
                             "long-term-trends.csv")) |> 
    mutate(parameter = case_when(parameter %in% c("chla_n", "nh4f", "no23f", "po4f", "turb_median") ~ paste0("log.", parameter),
                                 .default = parameter))

stns <- readr::read_csv(here::here("data_wq", "sampling_stations.csv")) |> 
    janitor::clean_names() |> 
    select(station = station_code, latitude, longitude) |> 
    mutate(longitude = as.numeric(longitude),
           longitude = longitude * -1)

wq_monthly <- left_join(wq_monthly, stns, by = "station")
```


What we learned from the national WQ synthesis is that we are seeing widespread increases in water temp and chl a, and widespread decreases in DO mg/L concentration. We are not however seeing increases in hypoxia itself (at least not yet). Maybe some increases in proportion of time DO is below 5 mg/L, though there are also some decreases; but no seeming changes (at the national scale) in proportion of time DO is below 2.   

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(trnds,
       aes(x = Slope,
           fill = sig_trend)) +
  geom_histogram(col = "gray40",
                 alpha = 0.8,
                 bins = 30) +
  facet_wrap(~parameter,
             scales = "free") +
  geom_vline(xintercept = 0,
             linewidth = 1) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Histograms of slope (change/yr) for each parameter",
       fill = "p < 0.05?")
```

What questions does this lead to?  

If we define hypoxia as "DO < 2 mg/L", we can ask some questions about how widespread it is:  

-  How many stations experienced [at least X months] with [at least X proportion] of hypoxic readings in the last year?  
-  How many stations experienced [at least X months] within a year, in at least [X] of the last 10 years?  
-  What is the spread of monthly values where hypoxia is logged at least [10%] of the time?  


Let's start with general spread of values for proportion of readings below 2 and 5, and the monthly level, across all stations:  

```{r}
ggplot(wq_monthly,
       aes(x = doLT2)) +
    geom_histogram(fill = "cadetblue3",
                   col = "gray60") +
    # scale_y_log10() +
    theme_bw()
```

Let's zoom in so it's not all dominated by all the 0s. The blue line is at 10%.   

```{r}
ggplot(wq_monthly,
       aes(x = doLT2)) +
    geom_histogram(fill = "cadetblue3",
                   col = "gray60") +
    coord_cartesian(ylim = c(0, 1250)) +
    geom_vline(xintercept = 0.1,
               col = "blue",
               linetype = "dashed") +
    theme_bw()
```

Let's summarize by .... month?  

```{r}
ggplot(wq_monthly,
       aes(x = doLT2)) +
    geom_histogram(fill = "cadetblue3",
                   col = "gray60") +
    facet_wrap(~month) +
    coord_cartesian(ylim = c(0, 200)) +
    geom_vline(xintercept = 0.1,
               col = "blue",
               linetype = "dashed") +
    theme_bw() 
```

Unsurprisingly, most of the readings occur in June-August, with non-ignorable amounts in May and September.  

I think 10% may be too high of a threshold to be interesting, so I'll take a look at how many stations experience at least 5% of the time being hypoxic in a given month....  

By year and month, how many stations are hypoxic at least 5% and 10% of the time?  

```{r}
hyp_df <- wq_monthly |> 
    summarize(.by = c(year, month),
              pct05 = sum(doLT2 >= 0.05, na.rm = TRUE),
              pct10 = sum(doLT2 >= 0.1, na.rm = TRUE),
              pct25 = sum(doLT2 >= 0.25, na.rm = TRUE),
              pct50 = sum(doLT2 >= 0.5, na.rm = TRUE)) |> 
    mutate(date = lubridate::ymd(paste(year, month, "01")))
```

```{r}
ggplot(hyp_df,
       aes(x = date)) +
    geom_line(aes(y = pct05,
              col = "at least 5% of the time")) +
    geom_line(aes(y = pct10,
              col = "at least 10% of the time")) +
    scale_color_brewer(palette = "Set1") +
    theme_bw() +
    labs(y = "# of sites experiencing hypoxia",
         col = "extent")
```

```{r}
ggplot(hyp_df,
       aes(x = as.factor(month),
           y = pct05)) +
    geom_boxplot()
```

Same pattern as before; more common in summer.  


What about those dips in certain years? Does it look like that in boxplot form too?  

```{r}
ggplot(hyp_df,
       aes(x = as.factor(year),
           y = pct05)) +
    geom_boxplot()
```

Not super informative, honestly.  

# Map 2: Summer hypoxia  

Maybe a good question to ask is, where are these stations that experience a lot of hypoxia in summer? Or, out of the 4 key months, how many of those months experience more than 5% hypoxic time at each station?  

```{r}
hyp_df_summ <- wq_monthly |> 
    filter(month %in% c(6, 7, 8, 9)) |> 
    summarize(.by = c(station, year),
              nMonths = sum(doLT2 >= 0.05, na.rm = TRUE)) |> 
    left_join(stns, by = "station")

hyp_df_summ2 <- hyp_df_summ |> 
    summarize(.by = station,
              nMonths = median(nMonths, na.rm = TRUE)) |> 
    left_join(stns, by = "station")
```

```{r}
ggplot(hyp_df_summ,
       aes(x = nMonths)) +
    geom_histogram()

ggplot(hyp_df_summ,
       aes(x = nMonths)) +
    geom_histogram() +
    coord_cartesian(ylim = c(0, 200))
```

```{r}
ggplot(hyp_df_summ2,
       aes(x = nMonths)) +
    geom_histogram() +
    ggtitle("Median # months with hypoxia, by station")

ggplot(hyp_df_summ2,
       aes(x = nMonths)) +
    geom_histogram() +
    ggtitle("Median # months with hypoxia, by station") +
    coord_cartesian(ylim = c(0, 15))
```


Or, if we can interactively choose a year..... and see how many stations had certain amounts of hypoxia that July-August.....  

```{r}
hyp_summer <- wq_monthly |> 
    filter(month %in% c(7, 8)) |> 
    summarize(.by = c(station, year),
              prop_LT2 = sum(doLessThan2_total, na.rm = TRUE) / sum(doLessThan2_nValid, na.rm = TRUE)) |> 
    left_join(stns, by = "station")
```

What sort of categorizing can we do? < 5% of the time; and then what bins?  

Just about 80% of the summer values are under 5%. So we're dividing up about 20% of the remainder into whatever categories we'd like.

```{r}
hist(hyp_summer$prop_LT2, breaks = 30, ylim = c(0, 200))

# how many do have it less than 5% of the time?
sum(hyp_summer$prop_LT2 < 0.05, na.rm = TRUE)
sum(hyp_summer$prop_LT2 < 0.05, na.rm = TRUE) / sum(!is.na(hyp_summer$prop_LT2))



hyp_summer2 <- hyp_summer |> 
    filter(prop_LT2 >= 0.05)
quantile(hyp_summer2$prop_LT2, probs = c(0.25, 0.5, 0.75, 1))    
```

Greater than 50% seems like an important delineation to me.... How much is that?  

```{r}
sum(hyp_summer2$prop_LT2 > 0.5, na.rm = TRUE) / sum(!is.na(hyp_summer2$prop_LT2))
```

About 7%.  

Maybe we do, under 5%; 5-25%; 25-50%; above 50?  

```{r}
hyp_summer_cat <- hyp_summer |> 
    mutate(category = case_when(prop_LT2 < 0.05 ~ "< 5%",
                                 prop_LT2 < 0.25 ~ "5-25%",
                                 prop_LT2 < 0.5 ~ "25-50%",
                                 .default = "> 50%"),
           category = factor(category, levels = c("< 5%",
                                                  "5-25%",
                                                  "25-50%",
                                                  "> 50%")))

save(hyp_summer_cat, file = here::here("data_wq", "summer_hypoxia.rds"))
```

```{r}
# library(leaflet)
# library(mapview)
```

```{r}
# palette <- colorFactor(palette = "YlOrRd", domain = hyp_summer_cat$category)
# 
# hyp_summer_cat |> 
#     filter(year == 2012) |> 
#     leaflet() |> 
#     addProviderTiles(leaflet::providers$Esri.WorldGrayCanvas) |> 
#     addCircleMarkers(lng = ~longitude,
#                      lat = ~latitude,
#                      color = ~palette(category),
#                      fillOpacity = 0.5,
#                      radius = 7) |> 
#     addLegend(position = "topright",
#               pal = palette,
#               values = ~category,
#               title = "Summer hypoxia",
#               opacity = 0.7)


```

